Chaotic mixing of tidal debris

### 2015-06-08 ms.tex
incorporating Kathryn's comments

\subsection{The behavior of chaotic orbits in non-integrable potentials}

@@

In previous work, we found that tidal debris from a simulated globular cluster (Palomar 5) on an orbit with a long Lyapunov time can form large ``fans'' of debris similar to those seen for progenitor systems on strongly chaotic orbits \citep{pearson15, fardal14}. Here, to illustrate the methods described above, we repeat the calculation of the Lyapunov exponent for the Pal 5, stream-fanning orbit of \cite{pearson15}. We use a longer integration time (equivalent to 10000 orbital periods) for the Lyapunov time calculation and additionally compute the frequency diffusion time.

Figure~\ref{fig:pal5} shows the finite-time Lyapunov exponent computed for the Pal 5 orbit --- the shaded region shows the ``converged'' part of the estimate that is used to compute the maximum Lyapunov exponent. By inverting the rate, we find $t_\lambda \approx 12~{\rm Gyr} \approx 40\,{\rm T}$, shorter than our finding in \cite{pearson15} (from a shorter integration time) but still longer (by a factor of two) than the N-body simulations that display stream fanning. This discrepancy highlights the difficulty in reliably computing the Lyapunov exponent as it can take many thousands of orbital periods to converge. We compute the frequency diffusion rate (Equation~\ref{eq:fdrate}) using two consecutive integration windows with length $\approx$50 orbital periods. From this, we compute a frequency diffusion time $t_\Omega = 35~{\rm T}$ by assuming $f=0.005$ (the spread in frequencies in a $10^4~\msun$ globular cluster on this orbit is $\approx$0.5\%). Though this orbit is computed in a different potential than will be used below, this demonstrates the apparent discrepancy between the timescales over which chaos is estimated to be relevant and the integration time over which stream-fanning becomes apparent. % Below, we [...] try to understand by looking at grids of orbits...

@@

% Figure 2
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/pal5-lyap.png}
\caption{ FTMLE (Equation~\ref{eq:ftmle}) computed over XXX orbital periods for the stream-fanning Palomar 5 orbit of \cite{pearson15}. The FTMLE saturates to the Lyapunov exponent in the shaded region and is estimated to be $\lambda_{\rm max} \approx XXX \, {\rm Gyr}^{-1}$, corresponding to a Lyapunov time of $t_\lambda \approx XX \, {Gyr}$. [TODO: add labels, remake with better font, remove blue line, add axis labels, add shaded region, check on all numbers for paragraph] } \label{fig:pal5}
\end{center}
\end{figure*}

% Figure 3
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/lyap_map.png}
\caption{ A grid of isoenergy orbits initialized on the $xz$ plane. Each pixel in this image represents a single orbit, and the greyscale corresponds to the logarithm of the Lyapunov time (in units of orbital periods). Chaotic orbits with $t_\lambda \gtrsim XX~{\rm T}$ appear regular because the integration time for each orbit is insufficient to resolve weak chaos. [TODO: remove Gyr and express in terms of orbital periods, add title, run Lyapunov grid for 5000 orbital periods instead of 1000?]} \label{fig:lyapmap} 
\end{center}
\end{figure*}

% Figure 4
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/fdiff_map.png}
\caption{The same grid of initial conditions as Figure~\ref{fig:lyapmap}, but now each pixel is colored by the logarithm of the frequency diffusion time (again in units of orbital periods). The frequencies are computed in two consecutive windows, each of which has length equal to 50 orbital periods. The frequencies are measured precisely so that small changes in the frequencies can be detected over just $\approx$10s of orbits. [TODO: remove Gyr and express in terms of orbital periods, add title] } \label{fig:freqdiff}
\end{center}
\end{figure*}

% Figure 5
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/nbody.png}
\caption{Particle positions from the final snapshots of N-body simulations of a globular-cluster-mass satellite disrupting in a triaxial NFW potential (Section~\ref{sec:potential}). Each satellite is evolved for 16 orbital periods, and begins and ends at apocenter. For visualization, the positions have been rotated so that the angular momentum of the progenitor at the final snapshot are aligned with the $z$-axis. Top row shows the positions projected onto the resulting $xy$ plane (the instantaneous orbital plane of the progenitor at the end of the simulation), bottom row shows the positions projected onto the $yz$ plane. From left to right, the columns indicate that the progenitor is on a regular (R), weakly chaotic (W), or strongly chaotic (S) orbit. Even on the weakly chaotic orbit --- where the predicted chaotic timescale is XX orbital periods --- the resulting tidal stream displays stream-fanning and the ``oldest'' debris is more diffuse than in the regular case. [TODO: suptitle 'N-body simulations'?]} \label{fig:nbodysims}
\end{center}
\end{figure*}

% Figure 6
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensembles.png}
\caption{Final particle positions after integrating unbound, globular-cluster-sized ensembles of orbits generated around each of the three N-body orbits (Figure~\ref{fig:nbodysims}). The ensembles each contain 1000 particles and begin at pericenter rather than apocenter because they are meant to represent debris disrupted during a single (the first) pericentric passage. The particle positions qualitatively match the morphology of the ``oldest'' debris from the N-body simulations, though there are many more particles in these ensembles than are disrupted during the first pericentric passage of the simulations. [TODO: suptitle 'Orbit ensembles'?] } \label{fig:ensembles}
\end{center}
\end{figure*}

>>>>

\subsection{The behavior of chaotic orbits in non-integrable potentials}\label{sec:behavior-chaotic}

@@

In Section~\ref{sec:results1}, we generate grids of orbits in the potential described above to map the orbit structure of the potential. We classify each orbit in terms of the strength of chaos along the orbit as computed using the methods of Section~\ref{sec:methods}. With this initial classification, in Section~\ref{sec:results2} we follow the evolution of ensembles of orbits generated around each orbit in the initial grid. We find that the configuration-space density of ensembles around weakly chaotic orbits evolve faster (e.g., in mean density) than expected given the timescales over which chaos is computed to be relevant for the parent orbits. In Section~\ref{sec:results3}, we explain this phenomenon in the context of how chaotic diffusion occurs (e.g., Section~\ref{sec:sec:behavior-chaotic}).

@@

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/lyap_map.png}
\caption{ A grid of isoenergy orbits initialized on the $xz$ plane. Each pixel in this image represents a single orbit, and the greyscale corresponds to the logarithm of the Lyapunov time (in units of orbital periods). Chaotic orbits with $t_\lambda \gtrsim XX~{\rm T}$ appear regular because the integration time for each orbit is insufficient to resolve weak chaos. [TODO: remove Gyr and express in terms of orbital periods, add title, run Lyapunov grid for 5000 orbital periods instead of 1000?]} \label{fig:lyapmap} 
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/fdiff_map.png}
\caption{The same grid of initial conditions as Figure~\ref{fig:lyapmap}, but now each pixel is colored by the logarithm of the frequency diffusion time (again in units of orbital periods). The frequencies are computed in two consecutive windows, each of which has length equal to 50 orbital periods. The frequencies are measured precisely so that small changes in the frequencies can be detected over just $\approx$10s of orbits. [TODO: remove Gyr and express in terms of orbital periods, add title] } \label{fig:freqdiff}
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/nbody.png}
\caption{Particle positions from the final snapshots of N-body simulations of a globular-cluster-mass satellite disrupting in a triaxial NFW potential (Section~\ref{sec:potential}). Each satellite is evolved for 16 orbital periods, and begins and ends at apocenter. For visualization, the positions have been rotated so that the angular momentum of the progenitor at the final snapshot are aligned with the $z$-axis. Top row shows the positions projected onto the resulting $xy$ plane (the instantaneous orbital plane of the progenitor at the end of the simulation), bottom row shows the positions projected onto the $yz$ plane. From left to right, the columns indicate that the progenitor is on a regular (R), weakly chaotic (W), or strongly chaotic (S) orbit. Even on the weakly chaotic orbit --- where the predicted chaotic timescale is XX orbital periods --- the resulting tidal stream displays stream-fanning and the ``oldest'' debris is more diffuse than in the regular case. [TODO: suptitle 'N-body simulations'?]} \label{fig:nbodysims}
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensembles.png}
\caption{Final particle positions after integrating unbound, globular-cluster-sized ensembles of orbits generated around each of the three N-body orbits (Figure~\ref{fig:nbodysims}). The ensembles each contain 1000 particles and begin at pericenter rather than apocenter because they are meant to represent debris disrupted during a single (the first) pericentric passage. The particle positions qualitatively match the morphology of the ``oldest'' debris from the N-body simulations, though there are many more particles in these ensembles than are disrupted during the first pericentric passage of the simulations. [TODO: suptitle 'Orbit ensembles'?] } \label{fig:ensembles}
\end{center}
\end{figure*}
